---
title: "Diversified Portfolios Versus The S&P500"
author: "by : Chris at Savvy Analytics"
output:
  html_document:
    df_print: paged
    code_folding: hide        
---

<span style="color:blue"><font size="4">Project Goal  </font></span><br>
Traditional investing advise suggests that diversifying your long term growth portfolio across stocks, bonds and international markets will lead to either favorable returns, favorable risk or both.

Let's take some simple portfolios and see if we can prove that theory with historical data.

```{r message=FALSE, warning=FALSE, include=FALSE}
# Clear Environment and load libraries
# Optional memory clear
rm(list=ls())
# Disable Scientific Notation in printing
options(scipen=999)

# Load libraries
QuietLoad <- function(library) {
  suppressWarnings(suppressPackageStartupMessages(
    library(library, character.only=TRUE)))
}

RunMode = "Windows"

if (RunMode == "Linux") {
  ArchiveFile = "/home/woolyadmin/R/StockMarketSectorPECycles.RData"
} else {
  ArchiveFile = "C:\\Users\\Chris.Woolery\\OneDrive - Republic Airways\\Documents\\RepoArchive\\SavvyDemoProjects\\R\\StockMarketSectorPECycles.RData"
}


```



```{r}

# Environment and Functions Section

QuietLoad('tidyverse')
QuietLoad('quantmod')
QuietLoad('lubridate')
QuietLoad('scales')
QuietLoad('janitor')
QuietLoad('zoo')
QuietLoad('kableExtra')
QuietLoad('ggrepel')
QuietLoad('DT')

RJETBlue = "#003365"

PrettyTable = function(TableObject, TableTitle) {
    TableObject %>%
      kable("html", escape = FALSE,
            caption = paste0('<p style="color:black; font-size:18px">',
            TableTitle,
            '</p>')) %>%
        kable_styling("striped",
                      bootstrap_options = c("hover", "condensed"),
                      full_width = TRUE) %>%
        row_spec(0, color = "white", background = RJETBlue) 
}

 
```

Note : This article, like most of mine on R-Pubs, is both instructional and investigative. You can re-create any of these charts and tables from the embedded R code. Click the “Code” button to view.

```{r}

# Data Load and Initial Data Definition Section


PortfolioInitialInvestment = 100000

PortfolioList = tibble::tribble(
        ~Portfolio, ~Symbol,                                  ~SymbolName, ~Weight,
"AM - All Money Mkt",      "VMFXX",                "Vanguard Federal Money Fund",       1,
"AB - All Bond",           "PIMIX",                         "Pimco Income Trust",       1,
"AS - All SP500 Index",    "SPY",                               "SP500 SPDR ETF",       1,
"DI - Div Intl 4U 2E 2A 2X", "SPY",                               "SP500 SPDR ETF",     0.4,
"DI - Div Intl 4U 2E 2A 2X", "VEUSX",                        "Vanguard Euro Stock",     0.2,
"DI - Div Intl 4U 2E 2A 2X", "VEMAX",                    "Vanguard Emerging Stock",     0.2,
"DI - Div Intl 4U 2E 2A 2X", "VPADX",                    "Vanguard Asia Pac Stock",     0.2,
"DU - Div US 6S 3B 1M",      "SPY",                               "SP500 SPDR ETF",     0.6,
"DU - Div US 6S 3B 1M",      "PIMIX",                         "Pimco Income Trust",     0.3,
"DU - Div US 6S 3B 1M",      "VMFXX",                "Vanguard Federal Money Fund",     0.1
  )

SecuritySymbols = PortfolioList$Symbol %>% unique()

# THe following code runs in a hidden block as it creates messsages that can't be supressed
# getSymbols(SecuritySymbols, 
#            src='yahoo', from = '1900-01-01', 
#            to = Sys.Date(), warnings = TRUE)


```

```{r include=FALSE}
# Faking VMFXX from 1 Year Treasury Returns
suppressWarnings({
suppressMessages({
getSymbols("DGS1", src = "FRED")
})
})

VMFXX = DGS1 %>%
  data.frame() %>%
  mutate(Col1 = 0, Col2 = 0, Col3 = 0, Col4 = 0, Col5 = 0,
         QuoteDate = as.Date(row.names(.), "%Y-%m-%d"),
         Elapsed =  difftime(QuoteDate, lag(QuoteDate, n=1), unit = "days") %>%
                      as.numeric()) %>%
  rename(CloseValue = DGS1) %>%
  relocate(CloseValue, .after = everything()) %>%
  fill(CloseValue, .direction = c("down")) %>%
  filter(!is.na(Elapsed)) %>%
  mutate(IntRate = (CloseValue / 36500) * Elapsed,
         TotalReturn = cumprod(1 + IntRate)) %>%
  select(Col1, Col2, Col3, Col4, Col5, TotalReturn)

suppressWarnings({
suppressMessages({
getSymbols(SecuritySymbols[! SecuritySymbols %in% c("VMFXX")], 
           src='yahoo', from = '1900-01-01', 
           to = Sys.Date(), warnings = TRUE)
})
})
```


```{r}

# Data transformation section

# Function to convert yahoo stock quotes to a data frame
StockToDF = function(XTSObject, StockName) {
  data.frame(XTSObject) %>%
      # Use Close Value = 4 or Adjusted Value = 6 to include dividends and splits
      rename(CloseValue = 6) %>%
      mutate(Symbol = StockName,
             QuoteDate = as.Date(row.names(.), "%Y-%m-%d"),
             WeekBeg = floor_date(QuoteDate, unit = "week")) %>%
      group_by(WeekBeg) %>%
      mutate(MaxDate = max(QuoteDate)) %>%
      ungroup() %>%
      filter(QuoteDate == MaxDate) %>%
      dplyr::select(Symbol, WeekBeg, CloseValue) %>%
      filter(!is.na(CloseValue))}

for (i in 1:length(SecuritySymbols)) {
  Temp = StockToDF(get(SecuritySymbols[i]), SecuritySymbols[i])
  if (i == 1){
    TRANSecurity = Temp
  } else {
    TRANSecurity = bind_rows(TRANSecurity, Temp)
  }
}

suppressWarnings({
suppressMessages({
rm(list = SecuritySymbols)
rm(DGS1)
})
})

suppressWarnings({
suppressMessages({
TRANPortfolio = PortfolioList %>%
  left_join(TRANSecurity, by = "Symbol") %>%
  # Get first date of each symbol within a portfolio
  group_by(Portfolio, Symbol) %>%
    mutate(SymbolFirstDate = min(WeekBeg),
           SymbolLastDate = max(WeekBeg)) %>%
    ungroup() %>%
  # Filter so that there is a common starting date for portfolio history with all stocks
  # represented
  group_by(Portfolio) %>%
    mutate(PortfolioFirstDate = max(SymbolFirstDate),
           PortfolioLastDate = min(SymbolLastDate)) %>%
  # All Portfolios Start Together
  ungroup() %>%
  filter(WeekBeg >= max(PortfolioFirstDate) &
         WeekBeg <= min(PortfolioLastDate)) %>%


  # Filter to a specific start date if desired
  #
  #filter(WeekBeg >= "2014-10-01") %>%
  #
  # Reset the first data of each stock so that we can have a beginning share count
  group_by(Portfolio, Symbol) %>%
    mutate(SymbolFirstDate = min(WeekBeg)) %>%
    mutate(SharesOwned = if_else(WeekBeg == SymbolFirstDate,
                                 (Weight * PortfolioInitialInvestment) / CloseValue,
                                  NA)) %>%
  fill(SharesOwned, .direction = "down") %>%
  mutate(SymbolValue = SharesOwned * CloseValue,
         SymbolReturn = if_else(lag(SymbolValue, n = 1, default = 0) == 0,
                                0,
                                SymbolValue / lag(SymbolValue, n = 1, default = 0) -1),
         SymbolReturn = cumsum(SymbolReturn))
})
})

AGGPortfolio = TRANPortfolio %>%
  group_by(Portfolio, WeekBeg) %>%
  summarize(SymbolValue = sum(SymbolValue), .groups = "drop") %>%
  group_by(Portfolio) %>%
  mutate(FinalValue = if_else(WeekBeg == max(WeekBeg), SymbolValue, NA),
         FinalCAGR = if_else(!is.na(FinalValue),
                             paste0(dollar(FinalValue / 1000, accuracy = .1), "K",
                              " ",
                              percent((FinalValue / PortfolioInitialInvestment)^
                                (1/as.numeric(difftime(max(WeekBeg), min(WeekBeg), units = "days") / 365.4))-1
                                ,
                                accuracy = 0.1)),
                      NA)) %>%
  ungroup() %>%
  mutate(Portfolio = factor(Portfolio))


ColorLevels <- levels(AGGPortfolio$Portfolio)
ColorVec <- setNames(hcl(h = seq(15, 375, length = length(ColorLevels)), 
                     l = 65, 
                     c = 100), 
                     ColorLevels)
ColorVec[c("AB - All Bond", "AM - All Money Mkt", "AS - All SP500 Index")] <- c("grey", "black", "red")

LineTypeVec = c(rep("solid", 3), rep("longdash", length(ColorLevels)-3))


```

Here are the simple portfolios I have created and their contents:

1. AS - All SP500 Index : The US stock market via the S&P 500 "SPY" ETF.
2. AB - All Bond : The US bond market via PIMCO Income fund "PIMIX".
3. AM - All Money Market : Money market rates via Vanguard Federal Money Market "VMFXX".
4. DU - Div US 6S 3B 1M : Diversified US with 60% SPY, 30% PMIX and 10% VMFXX.
5. DI - Div Intl 4U 2E 2A 2X : Diversified International Stock with 40% SPY, 20% Vanguard Euro "VEUSX", 20% Vanguard Asia Pacific "VPADX" and 20% Vanguard Emerging Market "VEMAX".

I could have used other funds but these are fairly generic, on the lower end in terms of fees and on the higher end in terms of total assets.
 
```{r}

datatable(PortfolioList,
          rownames = FALSE,
            caption = htmltools::tags$caption(
              "Portfolios and Contents",
              style =   "color: darkblue; 
                        font-size: 20px; 
                        text-align: center;"),
          filter = 'top', 
          options = list(pageLength = 10, autoWidth = TRUE))
```

Let's look at the performance since the inception of the youngest of these funds on `r min(AGGPortfolio$WeekBeg)`.

```{r fig.height=8, fig.width=10}

AGGPortfolio %>%
  ggplot(aes(WeekBeg, SymbolValue)) +
    geom_line(aes(color = Portfolio), linewidth = 2) +
    geom_label(aes(WeekBeg, FinalValue, 
                   label =  FinalCAGR), 
               fill = "white", na.rm = TRUE, size = 4) +
    facet_wrap(~Portfolio, nrow = 3) +
    scale_color_manual(values = ColorVec) +
    scale_y_continuous(labels = dollar_format(accuracy = 1),
                       expand = c(0.15, 0)) +
    scale_x_date(expand = c(0.2, 0)) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(title = "Portfolio Performance Comparison - Gross Value and CAGR",
         subtitle = paste0("Weeks Ending : ", min(AGGPortfolio$WeekBeg), " - ", max(AGGPortfolio$WeekBeg)),
         x = "",
         y = "Portfolio Value",
         caption = "Starting Investment of $100K")


```

Here we see the ending value and the Compounded Average Annual Growth Rate (CAGR) which is the annualized rate of return that would produce that same ending value.

It looks like the US S&P 500 outperformed everything else in terms of return but what about risk?

Let's look at the range of the portfolio value where each data point is a week across the life of the portfolio.

```{r fig.height=5, fig.width=10}

AGGPortfolio %>%
  ggplot(aes(Portfolio, SymbolValue)) +
    stat_boxplot(geom ='errorbar', width = 0.2) + 
    geom_boxplot(aes(fill = Portfolio), width = 0.3) +
    geom_jitter(color = "grey40", width = 0.1, size = 1.5, alpha = 0.3) +
    scale_fill_manual(values = ColorVec) +
    scale_y_continuous(labels = dollar_format(accuracy = 1)) +
    theme_classic() +
    theme(legend.position = "none") +
    labs(title = "Portfolio Performance Comparison - Gross Value Variance",
         subtitle = paste0("Weeks Ending : ", min(AGGPortfolio$WeekBeg), " - ", max(AGGPortfolio$WeekBeg)),
         x = "",
         y = "Portfolio Value",
         caption = "Starting Investment of $100K")


```

The risk profile of the money market is the lowest but contrary to traditional advice, we see similar "worst case" losses when comparing the S&P 500 to the International equity portfolio.  

Likewise, the bond and diversified US portfolios had favorable "worst case" losses but the were still below the 25th percentile loss of the S&P500.

```{r}

AGGPortfolioRate = AGGPortfolio %>%
  group_by(Portfolio) %>%
  mutate(ReturnRate = SymbolValue / lag(SymbolValue, n = 1, default = NA) - 1,
         ReturnRate = if_else(is.na(ReturnRate), 0, ReturnRate)) %>%
  ungroup()

AGGPortfolioRate_Wide = AGGPortfolioRate %>%
  pivot_wider(id_cols = c(WeekBeg), names_from = Portfolio, values_from = ReturnRate)

AGGPortfolio_Summary = AGGPortfolioRate %>%
  mutate(Year = year(WeekBeg)) %>%
  group_by(Portfolio, Year) %>%
  summarize(StartDate = min(WeekBeg),
            EndDate = max(WeekBeg),
            TotalReturn = sum(ReturnRate), .groups = "drop")

#LM_ComparePortfolio = "AM - All Money Mkt"
LM_ComparePortfolio = "AS - All SP500 Index"

i = 35
for (i in 1:nrow(AGGPortfolio_Summary)) {
  TempLM_x = AGGPortfolioRate$ReturnRate[AGGPortfolioRate$Portfolio == LM_ComparePortfolio
                                         & year(AGGPortfolioRate$WeekBeg) == AGGPortfolio_Summary$Year[i]]
  Ending_x = AGGPortfolioRate$SymbolValue[AGGPortfolioRate$Portfolio == LM_ComparePortfolio
                                         & year(AGGPortfolioRate$WeekBeg) == AGGPortfolio_Summary$Year[i]]
  Ending_x = Ending_x[length(Ending_x)]
  TempLM_y = AGGPortfolioRate$ReturnRate[AGGPortfolioRate$Portfolio == AGGPortfolio_Summary$Portfolio[i]
                                         & year(AGGPortfolioRate$WeekBeg) == AGGPortfolio_Summary$Year[i]]
  Ending_y = AGGPortfolioRate$SymbolValue[AGGPortfolioRate$Portfolio == AGGPortfolio_Summary$Portfolio[i]
                                         & year(AGGPortfolioRate$WeekBeg) == AGGPortfolio_Summary$Year[i]]
  Ending_y = Ending_y[length(Ending_y)]
  TempLM = lm(TempLM_y ~ TempLM_x)
  Temp = tibble(
    Portfolio = AGGPortfolio_Summary$Portfolio[i],
    Year = AGGPortfolio_Summary$Year[i],
    Alpha = Ending_y / Ending_x,
    Beta = TempLM$coefficients[2] %>% as.numeric(),
    SDRes = TempLM$residuals %>% sd())  
  if (i == 1){
    AGGPortfolio_LM = Temp
  } else {
    AGGPortfolio_LM = bind_rows(AGGPortfolio_LM, Temp)
  }
}

AGGPortfolio_Summary = AGGPortfolio_Summary %>%
  inner_join(AGGPortfolio_LM,
             by = c("Portfolio", "Year"))


```

The value-range analysis is somewhat captive to the start date of our portfolio in terms of "worst case" so let's instead do a true alpha / beta analysis analysis.  [Investopedia - Alpha and Beta](https://www.investopedia.com/ask/answers/102714/whats-difference-between-alpha-and-beta.asp)

```{r fig.height=7, fig.width=10}

AGGPortfolio_Summary %>%
  ggplot(aes(Beta, Alpha)) +
    geom_hline(aes(yintercept = 1), color = "red") +
    geom_vline(aes(xintercept = 1), color = "red") +
    geom_point(aes(color = Portfolio), size = 6) +
    scale_color_manual(values = ColorVec) +
    scale_x_continuous(breaks = seq(0, 1.25, 0.125)) +
    scale_y_continuous(breaks = seq(0.2, 1.8, 0.2)) +
    theme(legend.position = "top") +
    labs(title = "Portfolio Risk and Return",
         subtitle = paste0("Weeks Ending : ", min(AGGPortfolio$WeekBeg), " - ", max(AGGPortfolio$WeekBeg)),
         x = "Relative Risk vs SP500 (Beta)",
         y = "Relative Return vs SP500 (Alpha)")


```

Each data point is a calendar year in the life of the portfolio (see the appendix for a detailed table).  

Typically alpha and beta are made in comparison to the investment class benchmark index but for this analysis, all are compared to the S&P 500 which is represented by the single red dot.

Dots higher than 1 on the Y (Alpha) axis represent years where the return was higher than the S&P 500 while dots lower than 1 represent years where the return was lower.

Dots less than 1 on the X (Beta) axis represent years where the risk was lower than the S&P 500 while dots higher than 1 represent years where the risk was higher.

In the past years of history, we can see that bonds and our diversified US portfolio are the only portfolios that have out-performed the S&P 500 in a significant percentage of the years.

```{r fig.height=5, fig.width=10}

AGGPortfolio_Summary %>%
  ggplot(aes(Year, TotalReturn)) +
    geom_line(aes(color = Portfolio, linetype = Portfolio), linewidth = 1.5) +
#    geom_point(aes(color = Portfolio), size = 2) +
    scale_y_continuous(labels = percent_format(accuracy = 1)) +
    scale_color_manual(values = ColorVec) +
    scale_linetype_manual(values = LineTypeVec) +
    theme_classic() +
    theme(legend.position = "top") +
    labs(title = "Portfolio Performance Comparison",
         subtitle = paste0("Weeks Ending : ", min(AGGPortfolio$WeekBeg), " - ", max(AGGPortfolio$WeekBeg)),
         x = "",
         y = "Portfolio Annual Return")
```

Here you can see all of those data points on a timeline.  The bond out-performance was nominal and tended to be in years where rates were dropping.

Finally let's look at the funds inside a few of the diversified portfolios to see how they performed individually.

```{r fig.height=5, fig.width=10}
XRay = "DU - Div US 6S 3B 1M"
DisplayDF = TRANPortfolio %>%
  filter(!is.nan(SymbolReturn) &  Portfolio == XRay) %>%
  group_by(Symbol, SymbolName) %>%
  mutate(MaxDate = max(WeekBeg),
         FinalValue = if_else(MaxDate == WeekBeg, SymbolValue, NA),
         FinalCAGR = if_else(!is.na(FinalValue),
                             paste0(dollar(FinalValue / 1000, accuracy = .1), "K",
                              " ",
                              percent((FinalValue / (Weight * PortfolioInitialInvestment))^
                                (1/as.numeric(difftime(max(WeekBeg), min(WeekBeg), units = "days") / 365.4))-1
                                ,
                                accuracy = 0.1)),
                      NA)) %>%
  ungroup() %>%
  mutate(Symbol = factor(Symbol),
         SymbolName = factor(SymbolName))


DisplayDF %>%
  ggplot(aes(WeekBeg, SymbolValue)) +
    geom_line(aes(color = SymbolName), linewidth = 1.5) +
    geom_label_repel(aes(label = FinalCAGR, color = SymbolName), fill = "white", na.rm = TRUE, 
                     force = 10, min.segment.length = 0, show.legend = FALSE) +  
    scale_y_continuous(labels = dollar_format(accuracy = 1)) +
    scale_x_date(expand = c(0.1, 0)) +
    theme_classic() +
    theme(legend.position = "top") +
    labs(title = paste0("Portfolio X-Ray Comparison - Ending Value and CAGR - ", XRay),
         subtitle = paste0("Weeks Ending : ", min(AGGPortfolio$WeekBeg), " - ", max(AGGPortfolio$WeekBeg)),
         x = "",
         y = "Return Rate")
```

The bond portion of the portfolio had about 2/3 of the total return of the S&P500 and generally had muted peaks and troughs in comparison.

```{r fig.height=5, fig.width=10}

XRay = "DI - Div Intl 4U 2E 2A 2X"
DisplayDF = TRANPortfolio %>%
  filter(!is.nan(SymbolReturn) &  Portfolio == XRay) %>%
  group_by(Symbol, SymbolName) %>%
  mutate(MaxDate = max(WeekBeg),
         FinalValue = if_else(MaxDate == WeekBeg, SymbolValue, NA),
         FinalCAGR = if_else(!is.na(FinalValue),
                             paste0(dollar(FinalValue / 1000, accuracy = .1), "K",
                              " ",
                              percent((FinalValue / (Weight * PortfolioInitialInvestment))^
                                (1/as.numeric(difftime(max(WeekBeg), min(WeekBeg), units = "days") / 365.4))-1
                                ,
                                accuracy = 0.1)),
                      NA)) %>%
  ungroup() %>%
  mutate(Symbol = factor(Symbol),
         SymbolName = factor(SymbolName))


DisplayDF %>%
  ggplot(aes(WeekBeg, SymbolValue)) +
    geom_line(aes(color = SymbolName), linewidth = 1.5) +
    geom_label_repel(aes(label = FinalCAGR, color = SymbolName), fill = "white", na.rm = TRUE, 
                     force = 10, min.segment.length = 0, show.legend = FALSE) +  
    scale_y_continuous(labels = dollar_format(accuracy = 1)) +
    scale_x_date(expand = c(0.1, 0)) +
    theme_classic() +
    theme(legend.position = "top") +
    labs(title = paste0("Portfolio X-Ray Comparison - Ending Value and CAGR - ", XRay),
         subtitle = paste0("Weeks Ending : ", min(AGGPortfolio$WeekBeg), " - ", max(AGGPortfolio$WeekBeg)),
         x = "",
         y = "Return Rate")
```

The international stock funds all under-performed the US stock market and the US bond market.

<span style="color:blue"><font size="4">Conclusion  </font></span><br>
Our simple diversified portfolios historically under-performed the S&P 500 over the whole time period but did perform better in some years.  

- The diversified bond portfolio delivered higher returns than the S&P 500 in half of the years with less volatility.   
- International stock funds consistently under-performed and represented more volatility.

Therefore don't automatically assume that diversification will improve your portfolio.  Ask questions, do research, do the math and in all ways...

Be Savvy


<br>
<br>
<br>
<span style="color:blue"><font size="4">Appendix  </font></span><br>

```{r}
AGGPortfolio_Summary %>%
  mutate(`Total Return` = percent(TotalReturn, accuracy = 0.1),
         Beta = comma(Beta, accuracy = 0.001),
         `Alpha` = comma(Alpha, accuracy = 0.001),
         `Residual Std Dev` = comma(SDRes, accuracy = 0.001)) %>%
  arrange(Year, Portfolio) %>%
  select(Year, Portfolio, `Total Return`, `Alpha`, Beta, `Residual Std Dev`) %>%
datatable(rownames = FALSE,
            caption = htmltools::tags$caption(
              "Portfolio Risk and Return Table",
              style =   "color: darkblue; 
                        font-size: 20px; 
                        text-align: center;"),
          filter = 'top', 
          options = list(pageLength = 10, autoWidth = TRUE))

```


