---
title: "S&P 500 Analysis of Peaks vs Troughs"
author: "by : Chris at Savvy Analytics"
output:
  html_document:
    df_print: paged
    code_folding: hide        
---


<span style="color:blue"><font size="3">Goal  </font></span><br>
This is a simple exploration of stock market peaks and troughs with a comparison of the current market to historic market troughs. 

<span style="color:blue"><font size="3">Background  </font></span><br>
Theoretically, as long as there is a functional economy in the world the general stock market (represented by the S&P 500 in this analysis) will climb higher over the long term.  There are two reasons to expect this:

  1. The companies in the index are the target of spending by consumers, businesses and government so naturally their revenues will grow at a rate that represents the economy as a whole.

  2. Those companies who perform the best by turning revenues into profit will stay in the index, those who do not perform will drop out of the index.  Given this continual vetting, we can expect that the index will not just match but outperform the economy as a whole.
    
However the stock market does not grow at a perfectly steady linear rate.  Instead, it has a history of reaching new all time highs and then drops before gradually climbing to another new all time high.

Let's explore what some of those look like.

```{r message=FALSE, warning=FALSE, include=FALSE}
# Clear Environment and load libraries
# Optional memory clear
rm(list=ls())
# Disable Scientific Notation in printing
options(scipen=999)

# Load libraries
QuietLoad <- function(library) {
  suppressWarnings(suppressPackageStartupMessages(
    library(library, character.only=TRUE)))
}

QuietLoad('tidyverse')
QuietLoad('quantmod')
QuietLoad('lubridate')
QuietLoad('scales')
 
 
```

```{r message=FALSE, warning=FALSE}

# This is my standard load package to load up a dataframe with:
# S&P 500, treasury rates and common sector ETF's for analytics of various strategies

ArchiveFile = "/home/woolyadmin/R/SP500Strategy_PT1010.RData"

Offline = FALSE

if (Offline) {

load(ArchiveFile)  

 

} else {

# Load data using the QuantMod package to retrive historical data
# Treasury Bonds and rates are provided by the St. Louis Federal Reserve "FRED" system
  
StockNames = c("SP500", "Communication Services" ,"Consumer Discretionary" ,"Consumer Staples" ,
                "Energy" ,"Financials" ,"Health Care" ,
                "Industrials" ,"Materials" ,"Real Estate" ,
                "Technology" ,"Utilities")
StockSymbols = c("^GSPC", "XLC" ,"XLY" ,"XLP" ,
                "XLE" ,"XLF" ,"XLV" ,
                "XLI" ,"XLB" ,"XLRE" ,
                "XLK" ,"XLU")

BondNames = c("Frd Funds", "1Yr Treasury", "2Yr Treasury", 
                "5Yr Treasury", "10Yr Treasury", "20Yr Treasury",
                "30Yr Treasury")
BondSymbols = c("DFF", "DGS1","DGS2", 
            "DGS5", "DGS10", "DGS20", 
            "DGS30")


# Get the S&P from Yahoo
getSymbols(StockSymbols, src='yahoo', from = '1900-01-01', 
           to = Sys.Date(), warnings = TRUE)

# Call use QuantMod's API into FRED
getSymbols(BondSymbols, src = "FRED")

# Function to convert yahoo stock quotes to a data frame
StockToDF = function(XTSObject, StockName) {
  data.frame(XTSObject) %>%
      rename(CloseValue = 4) %>%
      mutate(SymbolName = StockName,
             QuoteDate = as.Date(row.names(.), "%Y-%m-%d"),
             WeekBeg = floor_date(QuoteDate, unit = "week")) %>%
      group_by(WeekBeg) %>%
      mutate(MaxDate = max(QuoteDate)) %>%
      ungroup() %>%
      filter(QuoteDate == MaxDate) %>%
      dplyr::select(SymbolName, WeekBeg, CloseValue) %>%
      filter(!is.na(CloseValue))}

# Function to convert FRED bond quotes to a data frame
BondToDF = function(XTSObject, BondName) {
  data.frame(XTSObject) %>%
    rename(CloseValue = 1) %>%
    mutate(SymbolName = BondName,
           QuoteDate = as.Date(row.names(.), "%Y-%m-%d"),
           CloseValue = CloseValue / 100,
           WeekBeg = floor_date(QuoteDate, unit = "week")) %>%
    group_by(SymbolName, WeekBeg) %>%
    summarise(CloseValue = mean(CloseValue, na.rm = TRUE),
              .groups = "drop")}

AllSymbolsDF = bind_rows(
  StockToDF(GSPC, "SP500"),
  StockToDF(XLC,"SecTelecom"),
  StockToDF(XLY,"SecDiscret"),
  StockToDF(XLP,"SecStaples"),
  StockToDF(XLE,"SecEnergy"),
  StockToDF(XLF,"SecFins"),
  StockToDF(XLV,"SecHealth"),
  StockToDF(XLI,"SecIndust"),
  StockToDF(XLB,"SecMaterl"),
  StockToDF(XLRE,"SecRealEst"),
  StockToDF(XLK,"SecTech"),
  StockToDF(XLU,"SecUtil"),
  BondToDF(DFF, "FedFunds"),
  BondToDF(DGS1, "Tres01Yr"),
  BondToDF(DGS2, "Tres02Yr"), 
  BondToDF(DGS5, "Tres05Yr"), 
  BondToDF(DGS10, "Tres10Yr"), 
  BondToDF(DGS20, "Tres20Yr"), 
  BondToDF(DGS30, "Tres30Yr"), 
)

AllSymbolDF_Pivot = AllSymbolsDF %>%
  pivot_wider(id_cols = WeekBeg, names_from = SymbolName, values_from = CloseValue) %>%
  filter(complete.cases(.))

save(list=c("AllSymbolsDF", "AllSymbolDF_Pivot"), file = ArchiveFile) 

}

```

```{r include=FALSE}

# PeaksvsTroughDF Function
# Allows you to test various peak and trough buying models
# Model Name : Whatever name you want to call the model
# PeakTarget Example 1 - 0.15 : Buy when market is 15% down from Prior AllTime High
# TroughTarget Example 1.4 : Sell when market is up 40% from Lowest trough since Prior AllTime High
PeaksvsTroughDF = function(ModelName, PeakTarget, TroughTarget) {
  SP500Wip = AllSymbolsDF %>%
    pivot_wider(id_cols = WeekBeg, names_from = SymbolName, values_from = CloseValue) %>%
    mutate(ModelName = ModelName) %>%
    dplyr::select(ModelName, WeekBeg, SP500) %>%
    mutate(SP500MA10 = rollmean(SP500, 10, fill = NA, align = c("right")),
           HighValue = as.numeric(NA),
           HighDate = as.Date(NA),
           VarToHigh = as.numeric(NA),
           LowValue = as.numeric(NA),
           LowDate = as.Date(NA),
           VarToLow = as.numeric(NA),
           Signal = as.character(NA),
           Position = as.character(NA),
           PositionShares = as.numeric(NA),
           PositionCash = as.numeric(NA),
           PositionValue = as.numeric(NA)) %>%
    filter(WeekBeg >= as.Date('1990-01-01'))
  
  i = 1
    CurrHighValue = SP500Wip$SP500[i]
    SP500Wip$HighValue[i] = CurrHighValue
    CurrHighDate = SP500Wip$WeekBeg[i]
    SP500Wip$HighDate[i] = CurrHighDate
    SP500Wip$VarToHigh[i] = SP500Wip$SP500[i] - SP500Wip$HighValue[i]
    CurrLowValue = SP500Wip$SP500[i]
    SP500Wip$LowValue[i] = CurrLowValue
    CurrLowDate = SP500Wip$WeekBeg[i]
    SP500Wip$LowDate[i] = CurrLowDate
    SP500Wip$VarToLow[i] = SP500Wip$SP500[i] - SP500Wip$LowValue[i]
    CurrSignal = "Hold"
    SP500Wip$Signal[i] = "Hold"
    if (ModelName == "BuyAndHold") {
        CurrPosition = "Shares"
        SP500Wip$Position[i] = "Shares"
        CurrPositionShares = 10000 / SP500Wip$SP500[i]
        SP500Wip$PositionShares[i] = CurrPositionShares
        CurrPositionCash = 0
        SP500Wip$PositionCash[i] = 0
      } else {
        CurrPosition = "Cash"
        SP500Wip$Position[i] = "Cash"
        CurrPositionShares = 0
        SP500Wip$PositionShares[i] = 0
        CurrPositionCash = 10000
        SP500Wip$PositionCash[i] = 10000
      }
    SP500Wip$PositionValue[i] = 10000
  for (i in 2:nrow(SP500Wip)) {
    SP500Wip$HighDate[i] = if_else(SP500Wip$SP500[i] > CurrHighValue, SP500Wip$WeekBeg[i], CurrHighDate)
    SP500Wip$HighValue[i] = if_else(SP500Wip$SP500[i] > CurrHighValue, SP500Wip$SP500[i], CurrHighValue)
    SP500Wip$VarToHigh[i] = SP500Wip$SP500[i] - SP500Wip$HighValue[i]
    CurrHighDate = SP500Wip$HighDate[i]
    CurrHighValue = SP500Wip$HighValue[i]
    # Low is since last high so reset if CurrHighDate = HighDate as well as if new low
    SP500Wip$LowDate[i] = if_else(SP500Wip$SP500[i] < CurrLowValue |
                                  CurrHighDate == SP500Wip$WeekBeg[i], SP500Wip$WeekBeg[i], CurrLowDate)
    SP500Wip$LowValue[i] = if_else(SP500Wip$SP500[i] < CurrLowValue |
                                  CurrHighDate == SP500Wip$WeekBeg[i], SP500Wip$SP500[i], CurrLowValue)
    SP500Wip$VarToLow[i] = SP500Wip$SP500[i] - SP500Wip$LowValue[i]
    CurrLowDate = SP500Wip$LowDate[i]
    CurrLowValue = SP500Wip$LowValue[i]
    # Set Signal, Set Position, Set PositionShares, Set PositionCash
    # Using PeakTarget and TroughTarget
    SP500Wip$Signal[i] = if_else(SP500Wip$SP500[i] / SP500Wip$LowValue[i] > TroughTarget, "Buy",
      if_else(SP500Wip$SP500[i] / SP500Wip$HighValue[i] < PeakTarget, "Sell", "Hold"))
    if (SP500Wip$Signal[i] == "Buy" & SP500Wip$Position[i-1] == "Cash") {
      SP500Wip$PositionShares[i] = SP500Wip$PositionCash[i-1] / SP500Wip$SP500[i]
      SP500Wip$PositionCash[i] = 0
      SP500Wip$Position[i] = "Shares"
    } else { 
      if (SP500Wip$Signal[i] == "Sell" & SP500Wip$Position[i-1] == "Shares") {
        SP500Wip$PositionCash[i] = SP500Wip$PositionShares[i-1] * SP500Wip$SP500[i]
        SP500Wip$PositionShares[i] = 0
        SP500Wip$Position[i] = "Cash"
      } else {
        SP500Wip$Position[i] = SP500Wip$Position[i-1]
        SP500Wip$PositionCash[i] = SP500Wip$PositionCash[i-1]
        SP500Wip$PositionShares[i] = SP500Wip$PositionShares[i-1]
      }  
    }
    SP500Wip$PositionValue[i] = if_else(SP500Wip$Position[i] == "Cash", SP500Wip$PositionCash[i],
                                        SP500Wip$SP500[i] * SP500Wip$PositionShares[i])
  }
  return(SP500Wip)
}

AllModelsDF =  
  bind_rows(
    PeaksvsTroughDF("BuyAndHold", .0001, 1000)
  )

AllModelsDF = AllModelsDF %>%
  group_by(ModelName) %>%
  mutate(MaxWeek = max(WeekBeg)) %>%
  mutate(EndingValue = if_else(WeekBeg == MaxWeek, PositionValue, as.numeric(NA))) %>%
  ungroup()


```

```{r}

# This custom function will plot out peaks, troughs and variances from the SP500 dataframe
# given a date range
PeakPlot = function(EndingDate, PlotTitle) {

  DisplayDF = AllModelsDF %>%
    filter(WeekBeg <= EndingDate & ModelName == "BuyAndHold")
  
  HighDate = max(DisplayDF$HighDate)
  StartDate = HighDate - months(12)
  HighValue = max(DisplayDF$HighValue)
  LowValue = DisplayDF$LowValue[length(DisplayDF$LowValue)]
  LowDate = DisplayDF$LowDate[length(DisplayDF$LowDate)]
  CurrentValue = DisplayDF$SP500[length(DisplayDF$SP500)]
  PeakLabel = paste0(percent((CurrentValue / HighValue) -1, accuracy = 2), " vs Peak")
  TroughLabel = if_else(CurrentValue / LowValue > 1,
                        paste0("+",percent((CurrentValue / LowValue) -1, accuracy = 2), " vs Trough"),
                        "in Trough")
  PeakvsTroughLabel = paste0(percent((LowValue / HighValue) -1, accuracy = 2), " Peak vs Trough")
  
  DisplayDF = DisplayDF %>%
    filter(WeekBeg >= StartDate)
  
  # About 30 positions are displayable
  BreakWeeks = paste0(max(1, round(nrow(DisplayDF) / 30, 1)), " weeks")
  
  #print(paste0("Low : ", LowValue, " High : ", HighValue, " Mean : ", mean(LowValue, HighValue)))
  
  Plot1 = DisplayDF %>%
    ggplot(aes(WeekBeg, SP500)) +
      geom_line(aes(color = "Weekly Closing Value"), size = 2) +
      geom_point(aes(WeekBeg, HighValue, color = "Peak to Date"), size = 2) +
      geom_point(aes(WeekBeg, LowValue, color = "Trough Since Last Peak"), size = 2) +
      scale_color_manual(name = "", values = c("Weekly Closing Value" = "black",
                                               "Peak to Date" = "green",
                                               "Trough Since Last Peak" = "red")) +
      scale_x_date(breaks = BreakWeeks) +
      scale_y_continuous(labels = comma_format(accuracy = 1)) +
      theme_minimal() +
      theme(legend.position = "bottom",
            axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
      labs(title = paste0("SP500 - ", PlotTitle),
           subtitle = paste0("By Week Through : ", floor_date(EndingDate, unit = "week")),
           caption = "Peaks = All Time Highs, Troughs are Lowest Points after a Peak",
           x = "", y = "S&P 500 Index Value")
  
  Plot1 +
    # Peak Line and Label
    annotate("segment", size = 1,
             x = HighDate, xend = EndingDate,
             y = HighValue, yend = HighValue,
             arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
    annotate("label",
             label = PeakLabel,
             x = HighDate + difftime(EndingDate, HighDate) / 2,
             y = HighValue) +
    # Peak to Trough Line and Label
    annotate("segment", size = 1,
             x = HighDate, xend = LowDate,
             y = mean(c(LowValue, HighValue)), yend = mean(c(LowValue, HighValue)),
             arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
    annotate("label",
             label = PeakvsTroughLabel,
             x = HighDate + difftime(LowDate, HighDate) / 2,
             y = mean(c(LowValue, HighValue))) +     
    # Trough Line and Label
    annotate("segment", size = 1,
             x = LowDate, xend = EndingDate,
             y = LowValue, yend = LowValue,
             arrow = arrow(ends = "both", angle = 90, length = unit(.2,"cm"))) +
    annotate("label",
             label = TroughLabel,
             x = LowDate + difftime(EndingDate, LowDate) / 2,
             y = LowValue) 
  

}
```

First a look at the most recent peak prior to the Federal Reserve rate increases targeting inflation.

```{r fig.width=10}

PeakPlot(Sys.Date(), "Most Recent - 2022 Inflation Shock")

```

<font size="3">Note : Data is automatically refreshed and is current as of: `r Sys.Date()`
</font>

Now let's look at some of the historical peaks and troughs and how those looked.

```{r fig.width=10}

PeakPlot(as.Date("2020-05-01"), "2020 Covid Panic")

```


```{r fig.width=10}

PeakPlot(as.Date("2019-02-01"), "2019 Trade War Panic")

```

```{r fig.width=10}

PeakPlot(as.Date("2009-06-01"), "Financial Sector Crisis 2008 - 2009")

```

```{r fig.width=10}

PeakPlot(as.Date("2002-04-01"), "Dot-Com Bubble")

```

```{r fig.width=10}

PeakPlot(as.Date("1991-01-01"), "Desert Shield & Storm")

```

Is there a reliable investment strategy you could base around this data?  I'll leave that up to you.

Whether you are in a peak or a trough.

Stay Savvy.

```{r}

```

